Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-07-25 extended_catalog/paper/extended_catalog.tex
lots of improvements

\begin{abstract}
We present a new method for analyzing spatially extended sources with
the Large Area Telescope (LAT), the primary science instrument on the
{\em Fermi Gamma-ray Space Telescope (Fermi)}. Measuring the extension
of a LAT source is important for uniquely identifying it.  We provide
a series of Monte Carlo studies to validate this tool and calculate
the LAT's detection threshold to spatially extended sources.  We then
apply this tool to test all sources in the two year source catalog for
extension. We present on the detection of ten spatially extended sources
in addition to the twelve spatially extended sources reported in the
second Fermi LAT Catalog.
\end{abstract}

@@

Using one year of all sky surveying data, the LAT published a catalog of
source that were significantly detected.  Many of these were associated
with a variety of source classes including active galactic nuclei,
pulsars, galaxies, and supernova remnants.  Many of these classes can be
spatially resolved when observed at other frequencies and could possibly
be spatially resolvable in \gev by the LAT.

The LAT collaboration has previously reported on the discovered
a variety of spatially extended including the supernova remnants (SNRs)
IC443 and W51C (\cite{ic443,w51c}),
the pulsar wind nebulae (PWNe) MSH 15-52 and Vela X
(\cite{msh1552,velax}), and the Large Magellanic
and the Small Magalenic Cloud (\cite{lmc,smc}).  There are other
source classes which could possibly be spatially resolved by the LAT
including galaxy clusters, dark matter, and possibly new physics.

Furthermore, a variety of spatially extended but otherwise unassociated
sources have been detected in the galactic plane at \tev energies using
Air Cherenkov detectors. Most prominent was a survey of the galactic plane
using the High Energy Stereoscopic System (H.E.S.S) which discovered 14
spatially extended sources (\cite{HESS_plane_survey}).

Being able to spatially resolve the \gev emission from a source is
important for several reasons. Because of the large point spread
function (PSF), there are often several possible source counterparts.
Finding a matching morphology can be used to uniquely identify the source.
Because of the largely varying PSF of the LAT, the spatial and spectral
information about a source do not nicely decouple. Therefore, a biased
spatial model will bias the spectral model of the source and studying
the extension of the source can correct for this.  Furthermore, correctly
modeling a source's extension is important for improving the overall model
of the sky and removing excess residuals.  These residuals manifested
themselves in the one year catalog as spurious point sources in the
region region surrounding the Large Magellanic Cloud (\cite{first_cat}).

@@

Morphological studies of sources in the \gev energy range
using the LAT are challenging because of the significantly 
energy-dependent PSF and because of systematic errors in the galactic diffuse
emission.  The LAT's PSF is determined at lower energies by multiple
scattering in the tracker and the 68\% containment radius of the PSF is
approximately 5.1\deg at 100 \mev  when averaged over instrument acceptance
and including photons which convert in either the thick or thin layers of
the tracker. The PSF approaches a limit given by the granularity of the
tracker channels and is 0.14\deg at 10 \gev (\cite{on_orbit_calibration}).
Therefore, the higher energy photons are significantly more important for
resolving a source but are typically limited by statistics. 
A detailed analysis procedure must be used to maximize our
sensitivity over all energies.

@@

This analysis is not feasible using the official LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center (\cite{fssc})} because \gtlike can only 
fit the spectral parameters of the model. The extension fitting would
have to be done as an external loop which would typically be
too time consuming. What has typically
been done to study an extended sources is to fix the position of the
source and develop a profile of the likelihood as a
function of extension.  This approach is not optimal because the
position and extension of the source should be simultaneously maximized
to find the best fit parameters and to correctly compute the statistical
significance of the detection.  Furthermore since this approach is
computationally intensive, no large scale Monte Carlo effort can
be run to validate it.

We developed an alternate approach to fit spatially extended
sources. A separate maximum likelihood fitting package called
\pointlike has been developed by the LAT collaboration and we have
added to it the ability to analyze spatially extended sources. \pointlike
is described in the one year catalog paper and in Matthew Kerr's Ph.D thesis
(\cite{first_cat,matthew_kerr_thesis}).

@@

The convolution of radially symmetric extended sources is optimized by
using a semi-analytic calculation.  The observed photon distribution
can be written as the convolution of the source shape ($I_\text{src}$)
with the psf
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A'.
\end{equation}
For the LAT, the PSF is parameterized as
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $\sigma$ and $\gamma$ are free parameters.  For radially symmetric
extended sources, the angular part of the integral can be done analytically
\begin{equation}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{equation}
This convolution formula is used by \pointlike because it is faster
and more closely reduces to the PSF when the source's size is 
small.

@@

We found that the {\em HESSE} errors on the extension of a source
were sometimes badly estimated.  On the other
hand, the {\em Simplex} algorithm was prohibitively slow and would often
fail to convergence. Our approach to calculating extension errors
is to fix the position of the source and varying just the extension
using the {\em Simplex} algorithm.  This approach
makes the approximation that the covariance between each spatial parameter
is small but not that the likelihood is Gaussian. This method 
is shown schematically in figure~\ref{extension_profile} which shows
the change in \loglikelihood when varying the extension of
the SNR IC443.
The localization error separately calculated by fixing the extension to the
best fit value and fitting an ellipse to the likelihood as a function
of the spatial parameters. This
elliptical localization error correctly accounts for the covariance
in the position of the source and this same me method is used to
calculate the localization error of all point sources in the first LAT catalog
(\cite{first_cat}).

\subsection{Gtlike Crosscheck}
\label{gtlike_crosscheck}

\pointlike is important for LAT analysis
that require many iterations such as source localization and extension
fitting.  On the other hand, we expect that the spectral parameters
fit by \gtlike to be slightly more precise.
Furthermore, because \gtlike is the standard likelihood analysis package,
it has been more extensively validated for spectral analysis.
For that reason, in the following analysis we took the best fit
position and extension found by \pointlike and
refit the spectral spectral parameters using \gtlike.
\gtlike also provides a second calculation of the likelihood which gives a
second measure of \ts and \tsext. We do not consider a source to be
significant unless there is agreement between the two methods.

We found good agreement between the two methods. For the new
extended source candidates, table~\ref{alt_diff_model_results} shows
a comparison of \ts and \tsext computed using \pointlike and \gtlike.
Unless explicitly mentioned, all \ts, \tsext, and spectral parameters
were calculated using \gtlike using \pointlike's positions and extensions.

@@

There is a degeneracy between a spatially extended source and multiple
point sources.  To assess the possibility of source confusion,
we developed a function in \pointlike to simultaneously fit 
the position of two point sources.
The code uses \minuit to fit the position of the two point
sources.  To avoid projection effects, the fit is done in a rotated coordinate
system.  To improve the robustness of the fit, Instead of directly fitting
the rotated coordinates $x_1$ and $x_2$ we instead
fit the sum and difference of the rotated
coordinates $S=(x_1+x_2)/2$ and $D=(x_2-x_1)/2$.  The \loglikelihood
is then maximized by fitting the sum and difference of the 
sources.  Because of \pointlike's computational efficiency, this algorithm
runs in a reasonable amount of time.

@@

Like extended sources described in section~\ref{gtlike_crosscheck},
the spectrum of the two point sources can be refit using \gtlike. 
we quote the spectral values obtained from \gtlike using the best fit positions
found using \pointlike.  For the extended candidates found by our search,
we computed both \tsext and \tsinc.  The results are presented in
table~\ref{dual_localization_results}.

@@

The spatial shape of extended sources can e modeled
as a two dimensional Gaussian
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{\pi\sigma^2}\delta\left(x^2+y^2<\sigma^2\right).
\end{equation}
Although these shapes are significantly different, the convolution
of them with the PSF for typical sizes look similar.  Even
though these shapes are significantly different, we demonstrate in
figure~\ref{compare_disk_gauss} that the convolution of shows the
PSF that would be observed for a power-law source of spectral index
2, the probability density function (PDF) for a Gaussian and uniform
disk of width $0.5\deg$, and the convolution of this PDF with the PSF.
The figure shows the convolution for the PSF in the energy range between
1 \gev and 100 \gev and also in the energy range between 10 \gev and
100 \gev.  The convolved shapes do separate better at higher energy, but
nevertheless this plot shows that we are not very sensitive to probing
the exact structure of an extended source.  In this paper we always
assume a uniform disk spatial model.

To get the convolved shapes to match in figure~\ref{compare_disk_gauss},
we had to set them to predict the same 68\% containment radius of 0.5\deg.  The
68\% containment radius for a spatial model is the radius at which 68\%
of the intensity is enclosed.  It can be computed by a straightforward
integral and we find that $\rsixeight_\text{,disk}=0.82\sigma$ whereas the
$\rsixeight_\text{,Gaussian}=1.51\sigma$.  Since $\rsixeight=0.5\deg$,
the uniform disk has $\sigma=0.41\deg$ whereas the Gaussian has
$\sigma=0.75\deg$.

This is important when comparing morphologies to other wavelengths
because the quoted size can vary significantly depending on which quantity
is reported.  This effect can be best seen in figure~\ref{compare_r68}
which a counts map of an extended source simulated in the energy
range from 1 \gev to 100 \gev with a uniform disk spatial model and
$\sigma=05$. Overlaid on the plot is the fit extension assuming the
source has a uniform disk and a Gaussian spatial model. The plot overlays
both $\sigma$ and $\rsixeight$ and $\sigma$ is significantly larger for
the disk spatial hypothesis compared to the Gaussian spatial model. On
the other hand, the disk and Gaussian's 68\% containment radius are
similar.

Since we are not very sensitive to the particular morphology of an
extended source, there is no correct choice for which size to quote
or which shape to overlay. For the following analysis, we quite the
disk's edge and overlay this size.  But due to smearing of the PSF,
$\sigma$ typically looks larger than the observed counts distribution.
When comparing the fit extensions with other wavelengths, it is important
to compare consistent quantities.

@@

We validated \pointlike's extension fitting code with a Monte Carlo
study. \tsext (defined in section~\ref{extension_fitting}) can be used to calculate the statistical significance
of the extension of a source.  Since the point source and extended source models are
nested, an extended source must always fit as well as a point source.  On the other hand, we would
not expect the likelihood to change very much unless the source was
significantly extended.  Since an extended source has one additional
degree of freedom, Wilks' Theorem says that the distribution of $\tsext$
when testing a point source for extension should follow a $\chi^2_1$
distribution (\cite{wilks_theorem}).

On the other hand, since the null hypothesis rests on the edge of
parameter space when the extension is 0, Wilks' Theorem does not hold
(\cite{warn_wilks_theorem}).  One might expect the actual distribution
to be a half $\chi^2_1$ distribution with one degree of freedom plus
half a delta function at 0:
\begin{equation}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
Chernoff proved that under certain assumptions this is the correct distribution
when the null hypothesis is on the edge of parameter
space (\cite{chernoff}).  The false detection
probability can be estimated by integrating this function from an observed
test statistic value to infinity. It is for this particular distribution
that the often quoted result holds that $\sqrt{\ts}$ is a measure of
the number of $\sigma$ of the detection (\cite{mattox_egret}).

@@

We simulated point sources of varying spectral models and fit them
them using \pointlike as both point and extended sources. The data was
simulated using \gtobssim\footnote{\gtobssim is distributed
by the Fermi Science Support Center (\cite{fssc}).}
The point sources were simulated with a power-law spectral model with
six $>100$ \mev fluxes ranging from $3\times 10^{-9} \ph/\cm^2/\sec$ to
$10^{-6} \ph/\cm^2/\sec$.  The point sources were simulated with spectral
indices ($\gamma$) of 1.5, 2, 2.5, and 3.  These values were picked to
to represent the types of sources detectable by the LAT.  The point
sources were simulated on top of an isotropic power-law background
with a Sreekumar-like spectrum ($>100$ \mev flux of $1.5\times 10^{-5}$ and spectral
index 2.1) (\cite{sreekumar_isotropic}).  The Monte Carlo simulation was
performed over a one year time interval using a default rocking profile
and a representative livetime fraction of 0.8.  The reconstruction was
performed with photons from an energy range of 1 \gev to 100 \gev using 4
energy bins per decade and using the 
Pass 7\_V6 (P7\_V6) Source Instrument
Response Function (IRFs).  For each point source, we used \pointlike
to fit it as both a point and extended source and calculate \tsext and
we only dept the sources which had a significant point source detection
($\ts>25$).

Before we present the results of the Monte Carlo simulations, we note
that despite out best efforts, there was always a small numerical
discrepancy between the likelihood when fitting the source as the PSF
and when fitting it as a very small extended source. This is due to
numerical error introduced by the convolution.  In most situations, this
error would be insignificant. But when testing a source for extension,
this offset significantly bias the distribution of \tsext in the null
hypothesis. To correct for this, instead of computing the likelihood in
the null hypothesis by fitting the PSF, we instead calculated \tsext by
fitting the sources as an extended source and as an extended source with
its extension fixed to ${10^{-10}}\deg$.

@@

We performed a second study to determine the LAT's detection
threshold to spatially extended sources. The detection threshold
to extension is defined as the flux at which the average
value of $\tsext$ is $\langle\ts\rangle=25$. As was seen in
section~\ref{monte_carlo_validation}, this corresponds to an extension
significance of $5\sigma$.  Qualitatively, we expect the LAT would not
be sensitive to sources much smaller than the PSF or much larger than
the PSF.  Small sources would look too much like the PSF and larger
sources would get lost in the background.

@@

Figure~\ref{index_sensitivity} shows the threshold for sources of four
spectral indices from 1.5 to 3 and extension varying from $\sigma=0.1\deg$
to $2.0\deg$.  The curves on the plot represent the flux for a given
extension where $\langle\tsext=25\rangle$.  The plot shows the threshold
using photons with energies between 100 \mev and 100 \gev and the
threshold using only photons with energies between 1 \gev and 100 \gev.
This results shows that the LAT's threshold to extension is better for
harder sources with more high energy photons. This result also shows
that the threshold is best for sources with an extension of $~0.5\deg$
and is significantly worse for very small extended small.

Except for the hardest and largest sources, our detection threshold
is almost unchanged when using events between 100 \mev and 1 \gev.
This is also demonstrated in figure~\ref{ts_ext_vs_energy} which shows
the \tsext for the SNR IC443 computed independently in 12 energy bins between 100 \mev and
100 \gev. For IC443, which has a spectral index ~2.4, almost the entire
increase in \ts comes from energies above 1 \gev.  On the other hand,
other systematic errors become increasingly important at low energy. For
our extension search, we use only only photons with energy above 1 \gev.

Figure~\ref{diff_factor_sensitivity} shows how the LAT's detection
threshold varies with increasing diffuse background. We compute the
sensitivity for a power-law source of spectral index 2 using 1, 10, and
100 times the extragalactic background. The latter factor approximately
represents the intensity of the galactic diffuse emission near the
galactic center.  The figure shows two plots, the first showing the
threshold using only photons with energies between 1 \gev and 100 \gev and
the second showing the threshold using only photons with energies between
10 \gev and 100 \gev. This plot demonstrates that we are less sensitive
to extended sources in a higher background region and that if the flux of
a source is constant (same number of photons), we would be more sensitive
to the extension of that source if those photons had a higher energy.

@@

The first and second LAT catalog produced a list of all significant point
sources seen by the LAT.  (\cite{first_cat,second_cat}).  The first LAT
catalog's source finding method assumes that all sources were consistent
with the PSF. The second LAT catalog modeled previously published
spatially extended sources with an extended spatial model but did not
attempt to resolve new spatially extended sources. In this paper,
we took the sources from the second Fermi LAT catalog and perform
a search to attempt to spatially resolve new sources.

For this search, we build a model of the
sky consistent with the second catalog.   We used the same two
years of data from August 4, 2008 to August 1, 2010 and we used the same
Pass 7\_V6 (P7\_V6) Source class event selections and 
Instrument Response Functions (IRFs, \cite{lat_on_orbit_psf}).  We used the same galactic diffuse,
isotropic diffuse, and earth limb emission model. The spectrum of the
galactic and isotropic diffuse models were refit during the analysis
while the earth limb emission was fixed to it's predicted value.

As was shown, we gain little in sensitivity using photons with energies
below 1 \gev. On the other hand, the large PSF at low energy makes us
more susceptible to systematic errors arising from source confusion due
to multiple point sources and susceptible to incorrect modeling of the
galactic diffuse emission.  For that reason, we performed our search
using photons only between 1 \gev and 100 \gev.
We used 4 energy bins per each logarithmic decade in energy.

>>>>

\begin{abstract}
We present a new method for analyzing spatially extended sources with
the Large Area Telescope (LAT), the primary science instrument on the
{\em Fermi Gamma-ray Space Telescope (Fermi)}. Measuring the extension
of a LAT source is important for uniquely identifying it.  We provide
a series of Monte Carlo studies to validate this tool and calculate
the LAT's detection threshold to spatially extended sources.  We then
apply this tool to test all sources in the two year source catalog for
extension. We present on the detection of nine spatially extended sources
in addition to the twelve spatially extended sources reported in the
second Fermi LAT Catalog (2FGL).
\end{abstract}

@@

Using two years of all sky surveying data, the LAT Collaboration
published a catalog of 1873 \gev sources.
Many of these were associated
with a variety of source classes including Active Galactic Nuclei (AGN),
pulsars, galaxies, and supernova remnants.  Many of these classes can be
spatially resolved when observed at other frequencies and could possibly
be spatially resolvable in \gev by the LAT.

The LAT collaboration has previously reported on the discovered of
five spatially extended supernova remnants (SNRs) IC443, W28, W44,
W51C, and RX J1713.7-2946 (\cite{ic443,w28,w44,w51c,rx_j1713_lat}).
The LAT Collaboration has reported on the detection of three extended
pulsar wind nebulae (PWNe) MSH 15-52 and Vela X, and HESS J1825-137
(\cite{msh1552,velax,fermi_hess_j1825}), the Large Magellanic
and Small Magalenic Cloud (\cite{lmc,smc}), and the radio galaxy
Centarus A (\cite{cen_a_lat}) .  There are other source classes which
could possibly be spatially resolved by the LAT including galaxy clusters
and dark matter satellite galaxies.

Furthermore, a variety of spatially extended but otherwise unassociated
sources have been detected in the Galactic plane at \tev energies using
Air Cherenkov detectors. Most prominent was a survey of the Galactic
plane using the High Energy Stereoscopic System (H.E.S.S) which presented
14 spatially extended sources (\cite{hess_plane_survey}) sources with
extensions varying from $~0.1\deg$ to $~0.25\deg$. Most \tev sources
in the galaxy are spatially extended and it is expected that these same
sources would have have extended \gev emission.

Being able to spatially resolve the \gev emission from a source is
important for several reasons. Because of the large point spread
function (PSF), there are often several possible source counterparts.
Finding a matching morphology can be used to uniquely identify the source.
Because of the largely varying PSF of the LAT, the spatial and spectral
information about a source do not nicely decouple. Therefore, a biased
spatial model will bias the spectral model of the source and studying
the extension of the source can correct for this.  Furthermore, correctly
modeling a source's extension is important for improving the overall model
of the sky and removing excess residuals.  These residuals manifested
themselves in 1FGL as spurious point sources in the region region
surrounding the Large Magellanic Cloud (\cite{first_cat}).  For this
reason, we are motivated to take the 1873 sources that were detected in
2FGL and apply a search for those sources which are spatially extended.

@@

Morphological studies of sources in the \gev energy range
using the LAT are challenging because of the significantly 
energy-dependent PSF and because of systematic errors in the Galactic diffuse
emission.  The LAT's PSF is determined at lower energies by multiple
scattering in the tracker and the 68\% containment radius of the PSF is
approximately 5.1\deg at 100 \mev  when averaged over instrument acceptance
and including photons which convert in either the thick or thin layers of
the tracker. The PSF approaches a limit given by the granularity of the
tracker channels and is 0.14\deg at 10 \gev (\cite{on_orbit_calibration}).
Therefore, the higher energy photons are significantly more important for
resolving a source but are typically limited by statistics. 
A detailed analysis procedure must be used to maximize our
sensitivity over all energies.

@@

This analysis is not feasible using the official LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center (\cite{fssc})} because \gtlike can only 
fit the spectral parameters of the model. The extension fitting would
have to be done as an external loop which would typically be
too time consuming. What has typically
been done to study an extended sources is to fix the position of the
source and develop a profile of the likelihood as a
function of extension.  This approach is not optimal because the
position and extension of the source should be simultaneously maximized
to find the best fit parameters and to correctly compute the statistical
significance of the detection.  Furthermore since this approach is
computationally intensive, no large-scale Monte Carlo effort can
be run to validate it.

We developed an alternative approach to fit spatially extended
sources. A separate maximum likelihood fitting package called
\pointlike has been developed by the LAT collaboration and we have
added to it the ability to analyze spatially extended sources. \pointlike
is described in 1FGL and in Matthew Kerr's Ph.D thesis
(\cite{first_cat,matthew_kerr_thesis}).

@@

The convolution of radially symmetric extended sources is optimized by
using a semi-analytic calculation.  The observed photon distribution
can be written as the convolution of the source shape ($I_\text{src}$)
with the psf
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A'.
\end{equation}
For the LAT, the PSF is parameterized by a {\em King function}
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $\sigma$ and $\gamma$ are free parameters
(\cite{matthew_kerr_thesis}).  For radially symmetric extended sources,
the angular part of the integral can be done analytically
\begin{equation}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{equation}
This convolution formula is used by \pointlike because it is faster
and more closely reduces to the PSF when the source's size is 
small.

There will always be a small numerical discrepancy between the PSF and
a very small extended source due to numerical error in the convolution.
In most situations, this error would be insignificant. But when testing
a source for extension, this offset can significantly bias the estimate
of the significance of the extension of a source.  To correct for this,
when testing a source for extension we compare the likelihood fitting
the source with an extended shape to the likelihood fitting it with its
extension fixed to ${10^{-10}}\deg$.

@@

We found that the {\em HESSE} errors on the extension of a source
were sometimes badly estimated.  On the other hand, the {\em Simplex}
algorithm was prohibitively slow and would often fail to convergence. Our
approach to calculating extension errors is to fix the position of the
source and varying just the extension using the {\em Simplex} algorithm.
This approach makes the approximation that the covariance between each
spatial parameter is small but not that the likelihood is Gaussian. This
method is shown schematically in figure~\ref{extension_profile} which
shows the change in \loglikelihood when varying the extension of the
SNR IC443.  The localization error is separately calculated by fixing
the extension and spectrum of the source and fitting a parabola to the
likelihood as a function of position.

\subsection{\gtlike Crosscheck}
\label{gtlike_crosscheck}

\pointlike is important for LAT analysis that require many iterations
such as source localization and extension fitting.  On the other hand,
because of the fewer approximations used in calculating the likelihood
we expect that the spectral parameters fit by \gtlike to be slightly
more precise.  Furthermore, because \gtlike is the standard likelihood
analysis package, it has been more extensively validated for spectral
analysis.  For that reason, in the following analysis  we use \pointlike
to determinate the position and extension of a source and then determine
the spectrum using \gtlike. Both \gtlike and \pointlike can estimate
the statistical significance of the extension of a source and we require
that both methods agree to claim a detection.

We found good agreement between the two methods. For the new extended
source candidates, table~\ref{alt_diff_model_results} will shows a
comparison of \ts and \tsext computed using \pointlike and \gtlike.
Unless explicitly mentioned, all \ts, \tsext, and spectral parameters
were calculated using \gtlike using \pointlike's positions and extensions.

@@

There is a degeneracy between a spatially extended source and multiple
point sources.  To assess the possibility of source confusion,
we developed a function in \pointlike to simultaneously fit the position
of two point sources.  To improve the robustness of the fit, the sum and
difference of the positions of the two sources are fit in a coordinates
rotated to the equator.  The \loglikelihood is then maximized by fitting
the sum and difference of the positions.

@@

Like extended sources described in section~\ref{gtlike_crosscheck},
the spectrum of the two point sources can be refit using \gtlike.
we quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.  For the extended candidates found
by our search, we computed both \tsext and \tsinc.  The results will be
presented in table~\ref{dual_localization_results}.

@@

The spatial shape of extended sources can be modeled
as a two dimensional Gaussian
\begin{equation}\label{gauss_pdf}
  \text{PDF}(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}\label{disk_pdf}
  \text{PDF}(x,y)=\tfrac{1}{\pi\sigma^2}\delta\left(x^2+y^2<\sigma^2\right).
\end{equation}
Although these shapes are significantly different, the convolution of them
with the PSF look similar.  Figure~\ref{compare_disk_gauss} demonstrates
this by showing the PSF, PDF, and convolution of the PDF with the PSF
for a uniform disk and Gaussian spatial model of size 0.5\deg.  This plot
shows that we are not very sensitive to probing the exact structure of an
extended source.  In the following we assume a uniform disk spatial model.

To get the convolved shapes to match in figure~\ref{compare_disk_gauss},
we had to set them to predict the same 68\% containment
radius of 0.5\deg.  The 68\% containment radius for a
spatial model is the radius at which 68\% of the intensity is
enclosed.  It can be computed by a straightforward integral
and we find that $\rsixeight_\text{,Gaussian}=1.51\sigma$
whereas $\rsixeight_\text{,disk}=0.82\sigma$ where $\sigma$
is defined in equation~\ref{disk_pdf} for the Gaussian source and
equation~\ref{gauss_pdf} for the disk source.  Since $\rsixeight=0.5\deg$,
the uniform disk has $\sigma=0.41\deg$ and the Gaussian has
$\sigma=0.75\deg$.

This is important when comparing morphologies to other wavelengths
because the quoted size can vary significantly depending on which quantity
is reported.  This effect can be best seen in figure~\ref{compare_r68}
which a counts map of an extended source simulated in the energy
range from 1 \gev to 100 \gev with a uniform disk spatial model with
an extension $\sigma$ of 0.5. Overlaid on the plot is the fit extension
assuming the source has a uniform disk and a Gaussian spatial model. The
plot overlays both $\sigma$ and $\rsixeight$ and $\sigma$ is significantly
larger for the disk spatial hypothesis compared to the Gaussian spatial
model. On the other hand, the disk and Gaussian's 68\% containment radius
are similar.

Since we are not very sensitive to the particular morphology of an
extended source, there is no correct choice for which size to quote
or which shape to overlay. For the following analysis, we quote the
disk's edge and overlay this size.  But due to smearing of the PSF,
$\sigma$ typically looks larger than the observed counts distribution.
When comparing the fit extensions with other wavelengths, it is important
to compare consistent quantities.

@@

We validated \pointlike's extension fitting code with a Monte Carlo
study. \tsext (defined in section~\ref{extension_fitting}) can be used to calculate the statistical significance
of the extension of a source.  Since the point source and extended source models are
nested, an extended source must always fit as well as a point source.  On the other hand, we would
not expect the likelihood to change very much unless the source was
significantly extended.  Since an extended source has one additional
degree of freedom, Wilks' Theorem predicts that the distribution of $\tsext$
when testing a point source for extension should follow a $\chi^2_1$
distribution (\cite{wilks_theorem}).

However, since the null hypothesis rests on the edge of
parameter space when the extension is 0, Wilks' Theorem does not hold
(\cite{warn_wilks_theorem}).  One might expect the actual distribution
to be a half $\chi^2_1$ distribution with one degree of freedom plus
half a delta function at 0:
\begin{equation}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
Chernoff proved that under certain assumptions this is the correct distribution
when the null hypothesis is on the edge of parameter
space (\cite{chernoff}).  The false detection
probability can be estimated by integrating this function from an observed
test statistic value to infinity. It is for this particular distribution
that the often quoted result holds that $\sqrt{\ts}$ is a measure of
the number of $\sigma$ of the detection (\cite{mattox_egret}).

@@

We simulated point sources of varying spectral models and fit them
them using \pointlike as both point and extended sources. The data was
simulated using \gtobssim\footnote{\gtobssim is distributed
by the Fermi Science Support Center (\cite{fssc}).}
The point sources were simulated with a power-law spectral model with
six $>100$ \mev fluxes ranging from $3\times 10^{-9} \ph/\cm^2/\sec$ to
$10^{-6} \ph/\cm^2/\sec$.  The point sources were simulated with spectral
indices ($\gamma$) of 1.5, 2, 2.5, and 3.  These values were picked to
to represent typical source paramters in LAT-detected sources. The point
sources were simulated on top of an isotropic power-law background
with a Sreekumar-like spectrum ($>100$ \mev flux of $1.5\times 10^{-5}$ and spectral
index 2.1) (\cite{sreekumar_isotropic}).  The Monte Carlo simulation was
performed over a one year time interval using a default rocking profile
and a representative livetime fraction of 0.8.  The reconstruction was
performed with photons from an energy range of 1 \gev to 100 \gev using 4
energy bins per decade and using the 
Pass 7\_V6 (P7\_V6) Source Instrument
Response Function (IRFs).  For each point source, we used \pointlike
to fit it as both a point and extended source and calculate \tsext and
we only kept the sources which had a significant point source detection
($\ts>25$).

@@

We performed a second study to determine the LAT's detection
threshold to spatially extended sources. The detection threshold
to extension is defined as the flux at which the average
value of $\tsext$ is $\langle\tsext\rangle=25$. As was seen in
section~\ref{monte_carlo_validation}, this corresponds to an extension
significance of $5\sigma$.  Qualitatively, we expect the LAT would not
be sensitive to sources much smaller than the PSF or much larger than
the PSF.  Small sources would look too much like the PSF and larger
sources would get lost in the background due to low surface brightness.

@@

Figure~\ref{index_sensitivity} shows the threshold for sources of four
spectral indices from 1.5 to 3 and extension varying from $\sigma=0.1\deg$
to $2.0\deg$.  The curves on the plot represent the flux for a given
extension where $\langle\tsext\rangle=25$. 
This results shows that the LAT's threshold to extension is better for
harder sources with more high energy photons. This result also shows
that the threshold is best for sources with an extension of $~0.5\deg$
and is significantly worse for very small extended small.

Figure~\ref{index_sensitivity} shows the threshold using photons with
energies between 100 \mev and 100 \gev and also only using photons
with energies between 1 \gev and 100 \gev.  Except for the hardest
and largest sources, our detection threshold is almost unchanged when
including photons with energies between 100 \mev and 1 \gev.  This is also
demonstrated in figure~\ref{ts_ext_vs_energy} which shows \tsext for the
SNR IC443 computed independently in twelve energy bins between 100 \mev and
100 \gev. For IC443, which has a spectral index ~2.4, almost the entire
increase in \ts comes from energies above 1 \gev.  On the other hand,
other systematic errors become increasingly important at low energy. For
our extension search, we use only only photons with energy above 1 \gev.

Figure~\ref{diff_factor_sensitivity} shows how the LAT's detection
threshold varies with increasing diffuse background. We compute the
sensitivity for a power-law source of spectral index 2 using 1, 10, and
100 times the extragalactic background. The latter factor approximately
represents the intensity of the Galactic diffuse emission near the
Galactic center.  The figure shows two plots, the first showing the
threshold using only photons with energies between 1 \gev and 100 \gev and
the second showing the threshold using only photons with energies between
10 \gev and 100 \gev. This plot demonstrates that we are less sensitive
to extended sources in a higher background region and that if the flux of
a source is constant (same number of photons), we would be more sensitive
to the extension of that source if those photons had a higher energy.

@@

2FGL is a list of 1873 \gev sources detected by the LAT.  2FGL included
twelve previously published spatially extended sources but did not
attempt to fit the extension of these sources. Other than these sources,
all 2FGL sources were modeled as point sources and 2FGL did not attempt to
resolve the extension of new sources.  In this paper, we tested all 2FGL
sources for extension to search for new spatially extended \gev sources.

For this search, we build a model of the
sky consistent with 2FGL.   We used the same two
years of data from August 4, 2008 to August 1, 2010 and we used the same
Pass 7\_V6 (P7\_V6) Source class event selections and 
Instrument Response Functions (IRFs, \cite{lat_on_orbit_psf}).  We used the same Galactic diffuse,
isotropic diffuse, and earth limb emission model. The spectrum of the
Galactic and isotropic diffuse models were refit during the analysis
while the earth limb emission was fixed to it's predicted value.

As was shown, we gain little in sensitivity using photons with energies
below 1 \gev. On the other hand, the large PSF at low energy makes us
more susceptible to systematic errors arising from source confusion due
to multiple point sources and susceptible to incorrect modeling of the
Galactic diffuse emission. In addition, the Galactic diffuse emission
is more pronounced at lower energies due to its steep energy spectrum.
For that reason, we performed our search using photons only between 1
\gev and 100 \gev.  We used 4 energy bins per each logarithmic decade
in energy.

We also performed a search for extended sources using photons only between
10 \gev and 100 \gev. Even though we are testing the same sources, this
approach is complimentary because the Galactic diffuse emission is even
less dominant above 10 \gev. Furthermore, source confusion becomes less of
a problem since most LAT sources are not significant at energies above 10
\gev.  Using higher energies, we might expect to detect harder sources in
more complicated regions. This is especially beneficial
for regions near pulsars which cutoff before 10 \gev
and was successfully used in detecting HESS J1825-137
and MSH 15-52 with the LAT (\cite{msh1552,fermi_hess_j1825}).

For each catalog source, we tested it for extension using \pointlike
assuming the source had a uniform radially symmetric disk spatial model.
We used a circular $10\deg$ region of interest centered on our source and
included all catalog sources within $15\deg$ of the source of interest.
We refit the spectral parameters of sources within $2\deg$ of the source.
We modeled each source with a power-law spectral model. This was reasonable because for a source to be significantly
extended, it would have to be fairly hard and fit reasonably well by
a power-law. Second, since we began the analysis at either 1 \gev or
10 \gev, we didn't expect there to be significant curvature over this
energy range.
